Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-12-06 draft_lm1.tex
'guaribating' ;0) the Methods

Also, since $R_t > 1$ indicates sustained transmission, one may be 
interested in computing the probability of this event.
This can be easily achieved by integrating~(\ref{eq:densityMantel}) over the 
appropriate interval.
By noting that
\begin{align}
\label{cumprobMantel}
Pr(R_t > 1) &= 1 - \int_0^1 f_P(r)dr \\
            &= 1- Pr(\theta_t < \frac{1}{2})
\end{align}
one can compute the desired probability while avoiding dealing with the density 
in~(\ref{eq:densityMantel}) directly.

\subsection*{Mathematical modeling} % The model

@@

As the epidemic progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined 
in~\ref{eq:Rtestimate}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
does not discriminate between them due to the lack of serotype-specific data.

\subsection*{Bayesian parameter estimation}

For the fitting procedure the incidence time series was divided into $J=13$ 
epidemic 
windows that corresponded to significant raises in incidence and normalized to 
lie on the $[0,1]$ interval.
Also, as the numerical integration routine needs to treat $\beta(t)$ as a
continuous function of time because of the variable step size, we used linear 
interpolation to obtain values of $R_t$ for any time point.

We take a Bayesian approach to the estimation of $S_0$ for each window.
For a given interval $j = \{ t_j^{\text{start}}, t_j^{\text{end}} \} $ we 
observe an incidence time series $\mathbf{Y_{j}}$.
We are thus interested in the posterior distribution
\begin{equation}
 \label{eq:S0post}
 p(S_{0j}|\mathbf{Y_{j}}) \propto l(\mathbf{Y_{j}}|S_{0j}, R_t, m,  \tau 
)\pi(S_{0j}) 
\end{equation}
The likelihood $l(\mathbf{Y_{j}}|\cdot)$ is assumed to be a Normal distribution 
with fixed variance $\sigma^2$.
In this estimation procedure we kept $R_t$ fixed at fixed at the posterior mean 
obtained as described in TextS1 and fixed $\tau = 1/7\: \text{days}^{-1}$.
% \subsubsection*{Priors}
To complete the inference, we need to specify prior distributions for the 
parameters of interest.
We place a flat $\text{Beta}(1, 1)$ prior on $S_{0j}\:\forall j$
% 
% \begin{align}
%  S_{0j} &\sim Beta(1, 1) \\
%        m &\sim \\
% \end{align}

% \subsubsection*{Posterior approximation}

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R convergence diagnostic~\cite{brooks1998}.

@@

\begin{equation}
\label{eq:AR}
A=\frac{cases}{Pop.\, at\, risk} 
\end{equation}

Based on the what have been discussed so far, we can rewrite (\ref{eq:AR}) for 
each epidemic $j$ as
\begin{equation}
\label{eq:AR2}
 A_{j}=\frac{\sum Y_j}{S_{0j}}
\end{equation}
where $S_{0j}$ is the number of susceptibles before each epidemic $j$, 
estimated as described above.

% Results and Discussion can be combined.
\section*{Results and Discussion}

>>>>

Also, since $R_t > 1$ indicates sustained transmission, one may be 
interested in computing the probability of this event.
This can be easily achieved by integrating~(\ref{eq:densityMantel}) over the 
appropriate interval.
By noting that
\begin{align}
\label{cumprobMantel}
Pr(R_t > 1) &= 1 - \int_0^1 f_P(r)dr \\
            &= 1- Pr(\theta_t < \frac{1}{2})
\end{align}
one can compute the desired probability while avoiding dealing with the density 
in~(\ref{eq:densityMantel}) directly.
s
\subsection*{Mathematical modeling} % The model

@@

As the epidemic progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined 
in~\ref{eq:Rtestimate}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modeling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$.

\subsection*{Bayesian parameter estimation}
We take a Bayesian approach to the estimation of $S_0$.
First the incidence time series was divided into $J=13$ 
epidemic windows that corresponded to significant raises in incidence and 
normalized to lie on the $[0,1]$ interval.
For a given interval $j = \{ t_j^{\text{start}}, t_j^{\text{end}} \} $ we 
observe an incidence time series $\mathbf{Y_{j}}$.
We are thus interested in the posterior distribution
\begin{equation}
 \label{eq:S0post}
 p(S_{0j}|\mathbf{Y_{j}}) \propto l(\mathbf{Y_{j}}|S_{0j}, R_t, m, \tau 
)\pi(S_{0j}) 
\end{equation}
The likelihood $l(\mathbf{Y_{j}}|\cdot)$ is assumed to be a Normal distribution 
with fixed variance $\sigma^2$.
In this estimation procedure we kept $R_t$ fixed at fixed at the posterior mean 
obtained as described above and fixed $\tau = 1/7\: \text{days}^{-1}$.
To complete the inference, we need to specify prior distributions for the 
parameters of interest.
We place a flat $\text{Beta}(1, 1)$ prior on $S_{0j}\:\forall j$

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at  
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
Also, as the numerical integration routine implemented within BIP needs 
$\beta(t)$ to be available at 
arbitrary values of $t$, i.e., as continuous function of time because of the 
variable step size, we used linear interpolation to obtain values of $R_t$ for 
any time point.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R convergence diagnostic~\cite{brooks1998}.

@@

\begin{equation}
\label{eq:AR}
A=\frac{\text{\# cases}}{\text{Population at risk}} 
\end{equation}

Based on the what has been discussed so far, we can rewrite (\ref{eq:AR}) for 
each epidemic $j$ as
\begin{equation}
\label{eq:AR2}
 A_{j}=\frac{\sum Y_j}{S_{0j}}
\end{equation}
where $S_{0j}$ is the number of susceptibles before each epidemic $j$, which we
estimated as described above.

\section*{Results and Discussion}
